SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2019-03-01 scipy-1.0/paper.tex

\subsection*{Current status}

At the time of writing, SciPy produces a new release
approximately every six months, and some of the major
feature highlights from the three years preceding
SciPy 1.0 are discussed in a technical improvements
section below. Some longer-term milestone events are
highlighted in Figure~\ref{fig:timeline}.

\begin{figure}[H]
\begin{verbatim}
    2001: first SciPy release
    2005: transition to NumPy
    2007: creation of scikits
    2008: spatial module and first Cython code added
    2010: move to a 6-monthly release cycle
    2011: development moves to GitHub
    2011: Python 3 support
    2012: add a sparse graph module and unified optimization interface
    2012: removal of maxentropy
    2013: continuous integration with TravisCI
    2015: adding Cython interface for BLAS/LAPACK and a benchmark suite
    2017: add a unified C API with LowLevelCallable; remove weave
    2017: SciPy 1.0 release
\end{verbatim}
\caption{Timeline}
\label{fig:timeline}
\end{figure}

@@

Python, Cython, Fortran, C and C++ are the programming languages used to
implement scientific algorithms in the SciPy library. An analysis of our 
codebase using the \texttt{linguist} library\cite{linguistref} provides a 
detailed breakdown as \% composition by programming language in 
SciPy (Figure~\ref{fig:linguist}).

@@

    \begin{tikzpicture}[font=\sffamily]
    \begin{axis}[height=6cm, width=10cm,
    ybar, ymax=60, ymin=0,bar width=1cm,
    ylabel={\% of Scipy codebase},
    ytick={0,20,...,60},
    axis lines*=left,
    axis y line shift=0pt,
    enlarge x limits=0.15,
    symbolic x coords={py,f,c,cy,cpp},
    xticklabels={Python,FORTRAN,C,Cython,C++},
    ticklabel style={align=center,font=\sffamily\small,
    /pgf/number format/assume math mode=true},
    xtick=data,
    nodes near coords,
    nodes near coords style={/pgf/number format/assume math mode=true}
    ]
    \addplot coordinates {(py,49.5) (f,25.6) (c,19.5) (cy, 3.0) (cpp,2.3)};
    \end{axis}
    \end{tikzpicture}
    \caption{The breakdown of programming languages used in the
             SciPy library determined using the linguist library.
    	 Small ($<0.5 \%$) amounts of TeX, Matlab, Shell,
    	 and Makefile are excluded for clarity and mostly
    	 provide supporting roles in tests, building, and
    	 documentation.}
    \label{fig:linguist}
\end{figure}

@@

\texttt{scipy.sparse} offers seven sparse matrix data structures,
also known as sparse formats. The most important ones are the row- 
and column-compressed formats (CSR and CSC, respectively). 
These offer fast major-axis indexing and fast matrix-vector multiplication,
and are used heavily throughout SciPy and dependent packages.

Over the last three years, our sparse matrix handling internals have been
rewritten and performance has been improved. Iterating over and slicing of CSC
and CSR matrices is now faster by up to 35\% (Figure~\ref{fig:sparse-iter}),
 and the coordinate (COO)/ diagonal (DIA) to CSR / CSC matrix format
conversions are now faster (Figure~\ref{fig:sparse-conv}). Importantly,
SuperLU\cite{superlu_ug99} was updated to version 5.2.1, enhancing the
low-level implementations leveraged by a subset of our \texttt{sparse}
offerings.

@@

@@

As of SciPy version 0.19, it is possible for users to wrap low-level functions
in a \texttt{scipy.LowLevelCallable} object that reduces the overhead of
calling compiled C functions directly from Python.  
Supported low-level functions include \texttt{PyCapsule} 
objects, \texttt{ctypes} function pointers, and \texttt{cffi} function pointers. 
Furthermore, it is even possible to generate a low-level callback function 
automatically from a Cython module using \texttt{scipy.LowLevelCallable.from\_cython}.

\subsubsection*{Cython bindings for BLAS, LAPACK, and special}

SciPy has provided special functions and leveraged BLAS and
LAPACK\cite{LAPACK} routines for many years. SciPy now additionally
includes Cython\cite{behnel2011cython} wrappers for
many BLAS and LAPACK routines (added in 2015) and the special functions 
provided in the \texttt{scipy.{\allowbreak}special} submodule (added in 2016).  
These Cython wrappers are available in the modules
\texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_blas},
\texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_lapack}, and
\texttt{scipy.{\allowbreak}special.{\allowbreak}cython\_special} respectively.
When writing algorithms in Cython, it is typically more efficient to call
directly into the libraries SciPy wraps rather than indirectly, using SciPy's
Python APIs.  These low-level interfaces for Cython can also be used outside of
the SciPy codebase to gain access to the functions in the wrapped libraries
while avoiding the overhead of Python function calls.  This can give
performance gains of one or two orders of magnitude for many use cases.

Developers can also use the low-level Cython interfaces without linking against
the wrapped libraries\cite{blas-lapack-wrappers-scipy-2015}.  This lets other
extensions avoid the complexity of finding and using the correct libraries.
Avoiding this complexity is especially important when wrapping libraries
written in Fortran.  Not only can these low-level wrappers be used without a
Fortran compiler, they can also be used without having to handle all the
different Fortran compiler ABIs and name mangling schemes.

@@

The \texttt{scipy.stats} module provides more than 100 probability
distributions and a framework to implement additional distributions. Given an
expression for a probability distribution function (pdf), the framework will
calculate moments, cumulative distribution function (cdf), differential
entropy, and more. By default, these properties are calculated by generic code, e.g.
numerical integration of the pdf to obtain the cdf, but the
generic code can be replaced by explicit formulas or more accurate/efficient
implementations when available. For example, most distributions have an explicit formula for
the cdf, so it is used instead of integrating the pdf. The distributions
framework also allows the user to draw random variates from the distribution.
Drawing random variates now accepts a \texttt{random\_state} parameter, which
is either a \texttt{numpy.random.RandomState} object or an integer used to
generate such an object, to provide fine-grained user-level control over the
random state of a workflow.  Currently about 80 continuous univariate
distributions, 12 discrete univariate distributions and 10 multivariate
distributions are implemented.  Key recent distributions added to
\texttt{scipy.stats} include the histogram-based distribution in
\texttt{scipy.stats.rv\_histogram} and the multinomial distribution in
\texttt{scipy.stats.multinomial}, the first discrete multivariate distribution 
available in SciPy.

@@

\subsubsection*{Test suite}
\label{sec:testsuite}

The SciPy test suite is orchestrated by a continuous integration matrix that
includes posix and Windows (32/64-bit) platforms managed by Travis CI and
AppVeyor, respectively. Our tests cover Python versions 2.7, 3.4, 3.5, 3.6, and
include code linting with \texttt{pyflakes} and \texttt{pycodestyle}. There are more than $13000$
unit tests in the test suite, which is written for usage with the pytest
runner, and with a 76.5 \% coverage of approximately $204000$ lines of code
holding steady for the last six months (Figure~\ref{fig:coverage}).  Some of
the historical components of the codebase would still benefit from increased
test coverage. Documentation for the code is automatically built and hosted by
the CircleCI service to facilitate evaluation of documentation changes /
integrity.  Our full test suite also passes with PyPy3, a just-in-time compiled
version of the Python language.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{static/coverage-chart}
\caption{\% coverage of the SciPy codebase by unit tests over the six months
preceding June 28, 2018 as reported by the \texttt{codecov} service. The
analysis of lines of source code covered is performed automatically in our
continuous integration suite using the \texttt{pytest-cov} and \texttt{gcov}
libraries.}
\label{fig:coverage}
\end{figure}

@@

In addition to ensuring that unit tests are passing, it is important to ensure that the
the performance of the SciPy codebase improves over time. Since February 2015, the 
performance of the SciPy codebase has been monitored with Airspeed Velocity
(\texttt{asv}\cite{asvref}) library. Figure~\ref{fig:asvbench}
shows example benchmark results for \texttt{scipy.spatial.cKDTree.query} over 
roughly nine years of project history.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{static/asv_time_query_ckdtree}
\caption{Airspeed velocity benchmarks for \texttt{scipy.spatial.cKDTree.query}. 
The results are based on
Python 2.7 performance on the master branch of the project using NumPy 1.8.2
and Cython versions 0.27.3, 0.21.1, and 0.18 (for improved backward
compatibility). Only the $L^2$ (Euclidean) norm is shown here, and to improve
backward compatibility and sampling of the benchmarks there was no application
of toroidal topology to the \texttt{KDTree} (i.e. the \texttt{boxsize} argument 
is $None$).}
\label{fig:asvbench}
\end{figure}

@@

The scientific Python ecosystem includes many examples
of domain-specific software libraries building on top
of SciPy features, and then returning to the base SciPy library
to suggest and even implement improvements to enable
more effective science applications. For example, there
are common contributors to the SciPy and astropy core
libraries\cite{astropy-2018}, and what works well for 
one of the codebases, infrastructures, or communities 
is often transferred in some form to the other. At the
codebase level, the \texttt{binned\_statistic} functionality
is one such cross-project contribution---it was initially
developed in an Astropy-affiliated package
and then placed in SciPy after the fact.

@@

SciPy has a strong developer community and a massive user base. GitHub traffic
metrics report roughly 20,000 unique visitors to the source website between May
14 and May 27, 2018, with 721 unique copies (``clones") of the codebase over
that roughly two-week period. The developer community consists of 610 unique
contributors of source code, with $>19,000$ commits accepted into the codebase
(GitHub page data).

>>>>

\subsection*{SciPy Today}
At the time of writing, the SciPy library consists of nearly 
600,000 lines of code organized in 16 subpackages. 
Over 85,000 GitHub repositories and almost 5000 packages depend
on SciPy. SciPy's GitHub repository attracts approximately 20,000 unique 
visitors every two weeks, and more than 100 people contribute to a 
new release every six months. Some of the major
feature highlights from the three years preceding
SciPy 1.0 are discussed in Section \ref{sec:technical_improvements}, 
and milestones in its history are highlighted in Figure~\ref{fig:timeline}.
% At the time of writing, SciPy produces a new release
% approximately every six months, and some of the major
% feature highlights from the three years preceding
% SciPy 1.0 are discussed in a technical improvements
% section below. Some longer-term milestone events are
% highlighted in Figure~\ref{fig:timeline}.

\begin{figure}[H]
\centering
\includegraphics[width=0.95\textwidth]{static/scipy_timeline}
\caption{Major milestones from SciPy's initial release in 2001 to
the release of SciPy 1.0 in 2017. Logos reprinted with permission.}
\label{fig:timeline}
\end{figure}

@@

Python, Cython, Fortran, C and C++ are the programming languages used to
implement scientific algorithms in the SciPy library. An analysis of our code
base using the \texttt{linguist} library\cite{linguistref} provides a 
detailed breakdown as \% composition by programming language in 
SciPy (Figure~\ref{fig:linguist}).

@@

    \begin{tikzpicture}[font=\sffamily]
    \begin{axis}[height=6cm, width=10cm,
    ybar, ymax=60, ymin=0,bar width=1cm,
    ylabel={\% of Scipy code base},
    ytick={0,20,...,60},
    axis lines*=left,
    axis y line shift=0pt,
    enlarge x limits=0.15,
    symbolic x coords={py,f,c,cy,cpp},
    xticklabels={Python,FORTRAN,C,Cython,C++},
    ticklabel style={align=center,font=\sffamily\small,
    /pgf/number format/assume math mode=true},
    xtick=data,
    nodes near coords,
    nodes near coords style={/pgf/number format/assume math mode=true}
    ]
    \addplot coordinates {(py,49.5) (f,25.6) (c,19.5) (cy, 3.0) (cpp,2.3)};
    \end{axis}
    \end{tikzpicture}
    \caption{The breakdown of programming languages used in the
             SciPy library determined using the linguist library.
    	 Small ($<0.5 \%$) amounts of TeX, Matlab, Shell,
    	 and Makefile are excluded for clarity and mostly
    	 provide supporting roles in tests, building, and
    	 documentation.}
    \label{fig:linguist}
\end{figure}

@@

@@

\texttt{scipy.sparse} offers seven sparse matrix data structures,
also known as sparse formats. The most important ones are the row- 
and column-compressed formats (CSR and CSC, respectively). 
These offer fast major-axis indexing and fast matrix-vector multiplication,
and are used heavily in scikit-learn.

Over the last three years our sparse matrix handling internals have been
rewritten and performance has been improved. Iterating over and slicing of CSC
and CSR matrices is now faster by up to 35\% (Figure~\ref{fig:sparse-iter}),
 and the COO / DIA to CSR / CSC matrix format
conversions are now faster (Figure~\ref{fig:sparse-conv}). Importantly,
SuperLU\cite{superlu_ug99} was updated to version 5.2.1, enhancing the
low-level implementations leveraged by a subset of our \texttt{sparse}
offerings.

@@

As of SciPy version 0.19, it is possible for users to wrap low-level functions
in a \texttt{scipy.LowLevelCallable()} object that reduces the overhead for
calling compiled C functions directly from Python.  Supported low-level
functions include \texttt{PyCapsule} objects, ctypes function pointers, and
cffi function pointers. The low level function signature must be consistent
with the expectations of the routine it is passed to. For example, the
documentation for \texttt{scipy.ndimage.generic\_filter} defines two acceptable
C callback function signatures that may be used to produce functions that
operate on each element of image data with low overhead. The C code may be
generated using numba or Cython, for example, as long as the function call
signature matches the specifications. Furthermore, it is even possible to
generate a low-level callback function automatically from a Cython module using
\texttt{scipy.LowLevelCallable.from\_cython}.

\subsection*{Cython bindings for BLAS, LAPACK, and special}

SciPy has provided special functions and leveraged BLAS and
LAPACK\cite{LAPACK} routines for many years. SciPy now additionally
includes Cython\cite{behnel2011cython} wrappers for:
many BLAS and LAPACK routines (added in 2015), and the special functions 
provided in the \texttt{scipy.{\allowbreak}special} submodule (added in 2016).  
These Cython wrappers are available in the modules
\texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_blas},
\texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_lapack}, and
\texttt{scipy.{\allowbreak}special.{\allowbreak}cython\_special} respectively.
When writing algorithms in Cython, it is typically more efficient to call
directly into the libraries SciPy wraps rather than indirectly, using SciPy's
Python APIs.  These low-level interfaces for Cython can also be used outside of
the SciPy codebase to gain access to the functions in the wrapped libraries
while avoiding the overhead of Python function calls.  This can give
performance gains of one or two orders of magnitude for many use cases.

Developers can also use the low-level Cython interfaces without linking against
the wrapped libraries\cite{blas-lapack-wrappers-scipy-2015}.  This lets other
extensions avoid the complexity of finding and using the correct libraries.
Avoiding this complexity is especially important when wrapping libraries
written in Fortran.  Not only can these low-level wrappers be used without a
Fortran compiler, they can also be used without having to handle all the
different Fortran compiler ABI's and name mangling schemes.

@@

The \texttt{scipy.stats} module provides more than 100 probability
distributions and a framework to implement additional distributions. Given an
expression for a probability distribution function (pdf), the framework will
calculate moments, cumulative distribution function (cdf), differential
entropy, and more. Those properties are calculated based on generic code, e.g.,
numerical integration of the pdf to obtain the cdf.  Where possible, the
generic code is replaced by explicit formulas or more accurate or efficient
implementations. For example, most distributions have an explicit formula for
the cdf, so it is used instead of integrating the pdf. The distributions
framework also allows the user to draw random variates from the distribution.
Drawing random variates now accepts a \texttt{random\_state} parameter, which
is either a \texttt{numpy.random.RandomState} object or an integer used to
generate such an object, to provide fine-grained user-level control over the
random state of a workflow.  Currently about 80 continuous univariate
distributions, 12 discrete univariate distributions and 10 multivariate
distributions are implemented.  Key recent distributions added to
\texttt{scipy.stats} include the histogram-based distribution in
\texttt{scipy.stats.rv\_histogram} and the multinomial distribution in
\texttt{scipy.stats.multinomial} (used, for example, in natural language
processing\cite{Griffiths5228}). \texttt{multinomial} is the first and so far
the only multivariate discrete distribution available in SciPy.

@@

@@

The airspeed velocity (asv\cite{asvref}) library enables benchmarking Python packages
over their lifetimes, and the performance of the SciPy code base was monitored
with asv starting in February of 2015 (PR \#4501). In addition to ensuring that
unit tests are passing (see Test suite section below), confirming 
that performance generally
remains constant or improves over the commit hash history of the project allows
us to objectively measure that our code base is improving, to empower
scientific applications.

Consider the asv benchmark results shown in Figure~\ref{fig:asvbench}, spanning
roughly nine years of project history. These demonstrate the gradual
performance improvements in a nearest-neighbor search through
\texttt{scipy.spatial.cKDTree.query()}, and can be run using:

\texttt{python run.py run -e -s 800 --bench "\textbackslash
btime\_query\textbackslash b" "02de46a546..b3ddb2c"},

where \texttt{run.py} is a light wrapper script around \texttt{asv}
that is packaged with SciPy.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{static/asv_time_query_ckdtree}
\caption{Airspeed velocity benchmarks for scipy.spatial.cKDTree.query()
over a roughly nine year commit history time frame. The results are based on
Python 2.7 performance on the master branch of the project using NumPy 1.8.2
and Cython versions 0.27.3, 0.21.1, and 0.18 (for improved backward
compatibility). Only the L2 (Euclidean) norm is shown here, and to improve
backward compatibility and sampling of the benchmarks there was no application
of toroidal topology to the KDTree (boxsize argument was ignored).}
\label{fig:asvbench}
\end{figure}

Any pull request can be compared against the \texttt{master} branch with the
command \texttt{asv continuous master new-feature}. This will provide a
benchmark against the master branch and the branch a new feature is implemented
in. More features are available in the \texttt{asv} documentation\cite{asvdocs}, including 
arguments to select which benchmarks to run.

\subsubsection*{Test suite}

The SciPy test suite is orchestrated by a continuous integration matrix that
includes posix and Windows (32/64-bit) platforms managed by Travis CI and
appveyor, respectively. Our tests cover Python versions 2.7, 3.4, 3.5, 3.6, and
include code linting with pyflakes and pycodestyle. There are more than $13000$
unit tests in the test suite, which is written for usage with the pytest
runner, and with a 76.5 \% coverage of approximately $204000$ lines of code
holding steady for the last six months (Figure~\ref{fig:coverage}).  Some of
the historical components of the code base would still benefit from increased
test coverage. Documentation for the code is automatically built and hosted by
the CircleCI service to facilitate evaluation of documentation changes /
integrity.  Our full test suite also passes with PyPy3, a just-in-time compiled
version of the Python language.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{static/coverage-chart}
\caption{\% coverage of the SciPy code base by unit tests over the six months
preceding June 28, 2018 as reported by the \texttt{codecov} service. The
analysis of lines of source code covered is performed automatically in our
continuous integration suite using the \texttt{pytest-cov} and \texttt{gcov}
libraries.}
\label{fig:coverage}
\end{figure}

@@

The scientific Python ecosystem includes many examples
of domain-specific software libraries building on top
of SciPy features, and then returning to the base SciPy library
to suggest and even implement improvements to enable
more effective science applications. For example, there
are common contributors to the SciPy and astropy core
libraries\cite{astropy-2018}, and what works well for 
one of the code bases, infrastructures, or communities 
is often transferred in some form to the other. At the code
base level, the \texttt{binned\_statistic} functionality
is one such cross-project contribution---it was initially
developed in an Astropy-affiliated package
and then placed in SciPy after the fact.

@@

SciPy has a strong developer community and a massive user base. GitHub traffic
metrics report roughly 20,000 unique visitors to the source website between May
14 and May 27, 2018, with 721 unique copies (``clones") of the code base over
that roughly two-week period. The developer community consists of 610 unique
contributors of source code, with $>19,000$ commits accepted into the code base
(GitHub page data).
